Comparing MCMC algorithms in Stochastic Volatility Models using Simulation Based Calibration

### 2023-10-19 manuscript/main.tex

\title{Comparing MCMC algorithms in Stochastic Volatility Models using Simulation Based Calibration}
\author{Benjamin Wee}
\date{October 2023}

\begin{document}

@@

\begin{abstract}

    This research uses the simulation-based calibration (SBC) method of \citet{talts2020validating} to analyse two commonly used, competing Markov chain Monte Carlo algorithms for the purpose of estimating a posterior distribution in the context of a stochastic volatility model. In particular, the bespoke ‘off-set mixture approximation’ algorithm proposed by \citet{kim1998stochastic} is explored together with a generic Hamiltonian Monte Carlo algorithm implemented through Stan \citep{stan}. The SBC analysis involves a simulation study to assess whether each sampling algorithm has the capacity to produce valid inference for the correctly specified model, while also characterising statistical efficiency through effective sample size. Results show that Stan's No-U-Turn sampler, an implementation of Hamiltonian Monte Carlo, produces a well-calibrated posterior estimate while the celebrated off-set mixture approach is less efficient and poorly calibrated, though model parameterisation also plays a role.  Limitations and restrictions of generality are discussed.

@@

\section{Introduction}
    Stochastic volatility (SV) models are used in financial econometrics to model and predict the behaviour of financial asset returns. Typically expressed as a non-Gaussian state space model where the conditional variance of the observed return is treated as a latent random variable \citep{hull1987pricing, chesney1989pricing}, such models are difficult to estimate since the likelihood is unavailable in closed form. With more parameters than observations,\footnote{We refer to both static unknowns and time-varying latent state variables as 'parameters'.} SV models are commonly analysed using Bayesian methods obtained via a Markov Chain Monte Carlo (MCMC) algorithm designed to sample from the high dimensional joint posterior distribution. The focus of this paper is to investigate the performance of two competing MCMC algorithms, each applied to a SV model, in terms of the validity and statistical efficiency of the their resulting inference. Performance of these algorithms will also be compared with different parameterisations of the SV model to determine the sensitivity of MCMC calibration to model specification.

    \citet{kim1998stochastic} propose a sampling strategy for the so-called 'vanilla' stochastic volatility through an off-set mixture approximation. Conditional on the mixture allocations and static parameters, the forward filter backward sampling algorithm \citep{carter1994gibbs, fruhwirth1995bayesian} is used to jointly sample the latent states while conditionally  conjugate prior distributions are also suggested.  Since then, there have been many advances in statistical computing and algorithm design. In particular, Hamiltonian Monte Carlo (HMC) has be proposed as a generic and efficient sampling method for MCMC. Notably, HMC has been shown to be particularly well-suited to complex, high dimensional probabilistic models. Importantly, the adoption of such new techniques have been made widely available through the development of various open source libraries such as the Stan programming language \citep{stan} and the PyMC library \citep{pymc2023}. As the development of new algorithms rapidly increase, so does the need to develop new methods to assess their output. Developments in statistical workflow are required to test new algorithms as well as compare computational strategies used to estimate increasingly complex models.

    This research analyses and compares the calibration and efficiency of MCMC used to estimate stochastic volatility models using a simulation study. By calibration we refer to a MCMC algorithm that returns correct posterior, on average, conditional on the correct model specification. The first algorithm is the No-U-turn sampler (NUTS) variant of Hamiltonian Monte Carlo (HMC) as implemented in the Stan programming language \citep{hoffman2014no, betancourt2017conceptual, stan}. The second algorithm replicates Kim Shephard and Chib's (KSC) Gaussian off-set mixture model which is sampled using the aforementioned Kalman Filter and Metropolis Hastings algorithm. The objective is to analyse the efficiency of the MCMC and their ability to return correct posterior estimates.

@@

    One particular challenge to comparing models is the parameterisation. Different parameterisations of a model, whilst the same mathematically, can be very different in terms of sampling difficulty \citep{neal2003slice}. In this work two different parameterisations of the stochastic volatility model are compared for each sampling method. An algorithm may be sensitive to different parameterisations of the same model which may impact MCMC performance as described in \citet{strickland2008parameterisation}. The key findings of this research are HMC applied on a re-parameterised SV model produces the most calibrated and efficient posterior estimates relative to KSC's MCMC algorithm on the approximate off-set mixture model. Additionally, MCMC algorithms may perform better or worse based on calibration and efficiency depending on the parameterisation of the model.

@@

\section{Research Context}
This section describes the stochastic volatility model to be analysed in this research and discusses two examples to motivate the research objectives. Section 2.1 outlines the model specificaion and priors, Section 2.2 explores challenges with inference and convergence diagnostics with no ground truth and Section 2.3 mentions the limitations and pitfalls of running only a single simulation.

\subsection{Stochastic Volatility}
    The model of interest is the discrete time, univariate stochastic volatility model estimated by \citet{kim1998stochastic} using a Bayesian framework. SV is expressed as a state space model which links the observed (logarithmic) return of a given asset over period t, denoted $y_t$, to its latent conditional variance, $h_{t+1}$. This model is given by the following expressions
    \begin{align}
    y_t =& \space \,\exp(h_t/2) \varepsilon_t, \\
    h_{t+1} =& \space \, \mu +\phi(h_t - \mu) + \sigma_{\eta} \eta_t,  \\
    h_1 \sim& \space \,\mathrm{N}\left(\mu, \frac{\sigma_{\eta}^2}{1-\phi^2}\right), \\
    \varepsilon_t \sim& \space \,\mathrm{N}(0,1), \\
    \eta_t \sim& \space \,\mathrm{N}(0,1).
    \end{align}
    The measurement equation (1) describes the conditional distribution of $y_t$ given $h_t$, where $h_t$ is the conditional log variance of $y_t$. $h_1$ is a draw from a stationary distribution and the state equation $h_{t+1}$ follows a stationary process governed by the auto-regressive parameter $\phi$ such that $|\phi|<1$. This auto-regressive parameter represents the persistence or stickiness of log volatility and the dispersion parameter $\sigma_{\eta}$ is the constant variance of the states. $\epsilon_t$ and $\eta_t$ are standard normal white noise shocks and are uncorrelated with each other.

    KSC assign certain independent prior distributions and hyperparameters to the static parameters, with conjugate priors for $\mu$ and $\sigma^2$, given by
    \begin{align}
    \mu \sim& \space \,\mathrm{N}(0, 10^2) \\
    \sigma_{\eta}^2 \sim& \space \, \mathrm{IG}(shape=5/2, scale=(0.01\times 5) / 2) \\
    \phi^{\ast} \sim& \space \,\mathrm{Beta}(20, 1.5) \\
    \phi =& \space \, 2\phi^{\ast} - 1
    \end{align}
    The prior distribution for $\phi$ is referred to as a stretched beta distribution, as it corresponds to a transformed beta random variable, $\phi^*$, with support (-1, 1).

@@

    Convergence diagnostics are useful for identifying when Bayesian computation fail on real data. However, confounding issues may arise when attempting to diagnose the cause of computational problems. Real data is generated from an unknown data generating process. That is, the true parameter and model are unobservable, and it is very likely that the model being fit is not the true model. Therefore, failed diagnostic checks could be the result of attempting to fit a misspecified model, issues with the MCMC algorithm, or both. Furthermore, different sampling strategies will result in different posterior estimates for the same model. Assuming no issues with the computation or the model specification, convergence diagnostics do not provide any guidance to the analyst regarding which posterior estimate is to be preferred since there is no ground truth.

    To illustrate the concerning issues, the posterior distribution associated with the stochastic volatility model in (1)-(5) and a set of real financial returns is estimated using two different MCMC algorithms. The data are demeaned daily (logarithmic) returns on the S\&P 500 Index over the period from 1 January 2023 - 10 September 2023 with sample size $T=171$. Figure \ref{fig:realdataex} displays the marginal posterior estimates for the static parameters $\mu$, $\phi$ and $\sigma^2$, with corresponding summary statistics in Table \ref{tab:realdata}. Overall the two methods are similar in skew and shape. However, closer inspections suggests that $\phi$ has a fatter left tail for the KSC sampler and $\sigma^2$ is also heavier in the right tails relative to HMC. This is evident in the quantiles with the 25th quantile for $\phi$ being smaller at 0.875 for KSC compared to 0.910 in HMC. Furthermore, HMC has taller peaks at their mode relative to KSC with a tighter spread. The uncertainty intervals created from each posterior sample would also be quite different and lead to different results. However, we can only conclude that there are differences between these methods, not which is correct.
        \begin{figure}[H]
        \centering
        \includegraphics[scale=0.1]{motivating_example/real_data_ex.png}
        \caption{\textbf{Posterior distribution for static parameters fit on the S\&P 500 Index.} Posterior samples from the HMC (blue) and KSC (red) MCMC samplers. Marginal posterior densities are presented for $\mu$, $\phi$ and $\sigma^2$. Note that each panel has their own scale for the x and y axis. The parameter $\phi$ falls}
        \label{fig:realdataex}
    \end{figure}

    % Other methods for model selection in this context (for example, out of sample predictive performance), but for parameter estimation it is unclear which one should be selected. 
    \begin{table}[H]
        \centering
        \begin{tabular}{|c|c|c|c|c|c|c|c|} \hline 
        Parameter&  MCMC&Min& q25&  Median& Mean & q75&Max\\ \hline 
        $\mu$&  KSC&-2.03 & -0.472 & -0.354 & -0.355 & -0.231 & 1.31 \\
     $\mu$&  HMC&-4.55 & -0.508 & -0.373 & -0.370 & -0.229 &2.86  \\\hline 
     $\phi$&  KSC&0.384 & 0.875 & 0.928 & 0.902 & 0.958 & 0.997 \\
     $\phi$&  HMC&0.438 & 0.910 & 0.953 & 0.929 & 0.977 &1.00  \\ \hline 
     $\sigma^2$&  KSC&0.00237 & 0.00891 & 0.0138 & 0.0178 & 0.0210 & 0.130 \\ 
     $\sigma^2$&  HMC&0.00209 & 0.00731 & 0.0105 & 0.0131 & 0.0159 &0.107 \\ \hline
        \end{tabular}
        \caption{\textbf{Summary statistics for HMC and KSC algorithms}. Both sets of estimates appear reasonable but it is unclear from estimates on real data which estimate is closer the the truth.}
        \label{tab:realdata}
    \end{table}

\subsection{Challenge 2: Limitations of a single simulation}
    Diagnosing problems with the approximate posterior from models fit to real data is difficult because we do not have a ground truth to compare with. A strategy around this is to evaluate a model and algorithm on simulated data. One approach is to simulate data from a generative model using known parameters. Then fit the same model on the simulated data and see if the true parameters can be recovered. This gives us the benefit of defining the true parameters of the data generating process to be estimated. If the model and algorithm cannot adequately capture the true parameter, then we cannot be confident that it will provide reliable estimates on real data.

    Simulation is particularly important in the SV model as it has latent parameters. \citet{gelman2020bayesian} discuss that simulation is the only way we can check inference on latent variables. This is critical for the SV model since the underlying framework is a state space model with latent log volatility parameters. Latent variables are unobserved in real data and are only estimated in the context of the model. Simulation gives control over the data generating process which reveals what the model can infer about the latent variables.

@@

    Figure 2 shows marginal posterior distributions given a single set of 1000 returns generated by the SV model. The true parameters are $\mu=-0.389$, $\sigma^2_{\nu}=0.0115$ and $\phi=0.547$ and the model is fit using Stan's implementation of HMC, the No-U-Turn sampler. The 95\% credible intervals for $\mu$ and $\sigma^2$ cover the true parameter. The true parameter for $\phi$ however, is in the tails and outside the interval. Such an analysis may incorrectly conclude that the model fails to adequately estimate the $\phi$ parameter. However, it may be the case that the posterior distribution is correctly calculated using this algorithm and the results may be due to the features of this specific simulated data set.

@@

\section{Methodology}
The methodology section describes the simulation study used validate the posterior estimates from MCMC algorithms and the metrics used to empirically measure calibration and efficiency. Section 3.1 discusses the procedure for running SBC and the key results for defining calibration. Section 3.2 defines the diagnostics and metrics for evaluating SBC experiment results.
    \subsection{Simulation Design}
        Simulation Based Calibration (SBC) checks the calibration of posterior estimates generated by MCMC algorithms. SBC is conducted by comparing the distribution of rank statistics to the uniform distribution which arises when an algorithm is correctly calibrated. For each $k$ of $K$ SBC iterations, one value for each parameter is drawn from the prior. For this parameter a data set is generated from the model. Then a sampling method (originally using Stan in \citet{talts2020validating}) is used to obtain posterior samples given this data and the prior. Then the estimated posteriors are compared to the true values by calculating the rank. Under a calibrated model, the $K$ rank statistics for each parameter should be uniformly distributed.

        To illustrate the SBC procedure, let $\theta$ be an arbitrary scalar parameter from a model and $y$ represent a data set. Start with a single parameter $\theta^{sim}$ drawn from the prior distribution:
        \begin{align}
        \theta^{sim} \sim \pi(\theta),
        \end{align}
        generate a data set $y^{sim}$ conditionally given $\theta^{sim}$ and the model of interest\footnote{Since all computation presumes a correctly specified model, nottation to condition on the model is suppressed.}:
        \begin{align}
        y^{sim} \sim \pi (y|\theta^{sim}).
        \end{align}
        Then take $L$ draws from the estimated posterior distribution, conditional on this data set, and generated by the MCMC algorithm under investigation (here, HMC or KSC):
        \begin{align}
        \{\theta_1,\dots , \theta_{B}\} \sim \pi (\theta | y^{sim}).
        \end{align}

        \citet{talts2020validating} prove that the sample of rank statistics for a given parameter, produced from replicate posteriors, follow a uniform distribution over the integers $\{0,1\dots,B\}$ if the posterior samples follow the same distribution as the prior on average. The rank statistic for any one-dimensional function of parameters $f:\Theta\rightarrow\ \mathbb{R}$ produced by the $k^{\mathrm{th}}$ replicate posterior sample, is defined as
        \begin{align}
        r^{(k)} = rank(\{f(\theta_1^{(k)}),\dots , f(\theta_{B}^{(k)})\}, f(\theta^{sim})) = \sum_{b=1}^{B}1[f(\theta_{b}^{(k)}) < f(\theta^{sim})].
        \end{align}

@@

        A key result is that the posterior averaged over the data and true parameters will equal the prior distribution. This is shown in the following expression:
        \begin{align}
        \pi(\theta) &= \int \int \pi(\theta|y^{sim}) \pi(y^{sim}|\theta^{sim}) \pi(\theta^{sim})d\theta^{sim} dy^{sim}, \\
        &= \int \int \pi(\theta|y^{sim}) \pi(y^{sim},\theta^{sim}) d\theta^{sim} dy^{sim}.
        \end{align}
        That is, the average posterior for $\theta$ over the SBC iterations should approximately equal to prior distribution, if the algorithm is calibrated. Therefore, any deviation from the prior distribution and the average posterior draws produced from replicated posterior distributions means that the sampling methodology is not producing the correct posterior, on average.

@@

            The shape of the distribution of rank statistics gives insight into how an MCMC may be miscalibrated. Specifically, it gives information about the average posterior deviates from the prior distribution of a given parameter. Figures \ref{fig:underestimation} and \ref{fig:underdispersed} are recreated from \citet{talts2020validating} for convenience. These display examples of posterior miscalibration and the corresponding non-uniform rank statistics.

@@

            Figure \ref{fig:underdispersed} has large inflated frequencies on both ends of the histogram resulting in a `U-shape'. \citet{talts2020validating} describe this bias as under-dispersion of the posterior relative to the prior distribution on average. The average posterior is too narrow (or overconfident) relative to the spread of the prior resulting in inflated frequency on both ends of the rank statistic distribution.

@@

            \begin{figure}[H]
                \centering
                \includegraphics[scale=0.07]{methodology/rhs.png}
                \caption{Left: Non uniform rank statistics with inflated frequency on the right side. Right: Average posterior distribution is biased to the left side of the prior distribution. Bias in rank statistics distribution comes from posterior samples underestimating the true parameter on average. Note that the two plots are independently produced for illustration only.}
                \label{fig:underestimation}
            \end{figure}

            \begin{figure}[H]
                \centering
                \includegraphics[scale=0.07]{methodology/underdispersed.png}
                \caption{Left: Non uniform rank statistics with inflated frequencies on both ends (`U-shape' distribution). Right: Overlayed posterior and prior distributions for the same arbitrary parameter. Non-Uniformity in ranks comes from under-dispersed posterior distribution relative to prior.}
                \label{fig:underdispersed}
            \end{figure}

            \subsubsection{Chi squared statistics}
            One difficulty of relying on the visualisation of rank statistic distributions for SV models, is there are a very large number of unknowns in the model to consider. An alternative to visually checking for uniformity of a given parameter is to calculate a chi-squared statistic based on the counts in each histogram bin. Suppose the histogram has $J$ bins, where for each $j^{th}$ bin, $b_j$ is the number of counts and $e_j$ the expected count in bin $j$, where the expected count is the ratio of K rank statistics and J bins, $e_j=\frac{K}{J}$. The corresponding chi- squared statistic is given by:
            \begin{align}
            \chi^2 = \sum_{j=1}^J \frac{(b_{j} - e_{j})^2}{e_j}.\end{align}

>>>>

\title{Comparing MCMC algorithms in Stochastic Volatility Models using Simulation Based Calibration}
\author{Benjamin Wee}
\date{\small October 2023\\[0.5cm]{\small Supervised by Professor Catherine Forbes and Dr Lauren Kennedy}}
% \author{It's me\\[1cm]{\small Advisor: A.D.Visor}}
\begin{document}

@@

\begin{abstract}
    Simulation based calibration (SBC) \citep{talts2020validating} is applied to analyse two commonly used, competing Markov chain Monte Carlo algorithms for estimating the posterior distribution of a stochastic volatility model. In particular, the bespoke ‘off-set mixture approximation’ algorithm proposed by \citet{kim1998stochastic} is explored together with a Hamiltonian Monte Carlo algorithm implemented through Stan \citep{stan}. The SBC analysis involves a simulation study to assess whether each sampling algorithm has the capacity to produce valid inference for the correctly specified model, while also characterising statistical efficiency through effective sample size. Results show that Stan's No-U-Turn sampler, an implementation of Hamiltonian Monte Carlo, produces a well-calibrated posterior estimate while the celebrated off-set mixture approach is less efficient and poorly calibrated, though model parameterisation also plays a role. Limitations and restrictions of generality are discussed.

    % This research uses the simulation-based calibration (SBC) method of \citet{talts2020validating} to analyse two commonly used, competing Markov chain Monte Carlo algorithms for the purpose of estimating a posterior distribution in the context of a stochastic volatility model. In particular, the bespoke ‘off-set mixture approximation’ algorithm proposed by \citet{kim1998stochastic} is explored together with a generic Hamiltonian Monte Carlo algorithm implemented through Stan \citep{stan}. The SBC analysis involves a simulation study to assess whether each sampling algorithm has the capacity to produce valid inference for the correctly specified model, while also characterising statistical efficiency through effective sample size. Results show that Stan's No-U-Turn sampler, an implementation of Hamiltonian Monte Carlo, produces a well-calibrated posterior estimate while the celebrated off-set mixture approach is less efficient and poorly calibrated, though model parameterisation also plays a role. Limitations and restrictions of generality are discussed.

@@

\section{Introduction}
    Stochastic volatility (SV) models are used in financial econometrics to model and predict the behaviour of financial asset returns. Typically expressed as a non-Gaussian state space model where the conditional variance of the observed return is treated as a latent random variable \citep{hull1987pricing, chesney1989pricing}, such models are difficult to estimate since the likelihood is unavailable in closed form. With more parameters than observations\footnote{We refer to both static unknowns and time-varying latent state variables as 'parameters'.}, SV models are commonly analysed using Bayesian methods obtained via a Markov Chain Monte Carlo (MCMC) algorithm designed to sample from the high dimensional joint posterior distribution. The focus of this paper is to investigate the performance of two competing MCMC algorithms, each applied to a SV model, in terms of the validity and statistical efficiency of the their resulting inference. Performance of these algorithms will also be compared with different parameterisations of the SV model to determine the sensitivity of MCMC calibration to model specification.

    \citet{kim1998stochastic} propose a sampling strategy for the so-called `vanilla' stochastic volatility through an off-set mixture approximation. Conditional on the mixture allocations and static parameters, the forward filter backward sampling algorithm \citep{carter1994gibbs, fruhwirth1995bayesian} is used to jointly sample the latent states while conditionally  conjugate prior distributions are also suggested.  Since then, there have been many advances in statistical computing and algorithm design. In particular, Hamiltonian Monte Carlo (HMC) has been proposed as a generic and efficient sampling method for MCMC. Importantly, the adoption of such new techniques have been made widely available through the development of various open source libraries such as the Stan programming language \citep{stan} and the PyMC library \citep{pymc2023}. As the development of new algorithms rapidly increase, so does the need to develop new methods to assess their output. Developments in statistical workflow are required to test new algorithms as well as compare computational strategies used to estimate increasingly complex models.

    This research analyses and compares the calibration and efficiency of MCMC used to estimate stochastic volatility models using a simulation study. By calibration we refer to a MCMC algorithm that returns correct posterior, on average, conditional on the correct model specification. The first algorithm is the No-U-turn sampler (NUTS) variant of Hamiltonian Monte Carlo (HMC) as implemented in the Stan programming language \citep{hoffman2014no, betancourt2017conceptual, stan}. The second algorithm replicates Kim Shephard and Chib's (KSC) Gaussian off-set mixture model which is sampled using the aforementioned Kalman Filter and Metropolis Hastings algorithm. The objective is to analyse the efficiency of the MCMC and their ability to return correct posterior estimates on average.

@@

    One particular challenge to comparing models is the parameterisation. Different parameterisations of a model, whilst mathematically the same, can be very different in terms of sampling difficulty \citep{neal2003slice}. In this work two different parameterisations of the stochastic volatility model are compared for each sampling method. An algorithm may be sensitive to different parameterisations of the same model which may impact MCMC performance as described in \citet{strickland2008parameterisation}.

    The key findings of this research are HMC produces more calibrated and efficient posterior estimates relative to KSC's MCMC algorithm on the approximate off-set mixture model. Additionally, MCMC algorithms may perform better or worse based on calibration and efficiency depending on the parameterisation of the model. HMC applied on a re-parameterised SV model produces the most calibrated and efficient posterior estimates across all SBC experiments.

@@

\section{Research Context}
This section describes the stochastic volatility model and discusses two examples to motivate the research objectives. Section 2.1 outlines the model specification and priors, Section 2.2 explores challenges with inference and convergence diagnostics when the true model is unknown (no ground truth) and Section 2.3 discusses the limitations and pitfalls of running only a single simulation. Section 2.4 lays out the research objective given these motivating examples.

\subsection{Stochastic Volatility}
    The model of interest is the discrete time, univariate stochastic volatility model estimated by \citet{kim1998stochastic} using a Bayesian framework. SV is expressed as a state space model which links the observed (logarithmic) return of a given asset over period $t$, denoted $y_t$, to its latent conditional variance, $h_{t+1}$. This model is given by the following expressions
    \begin{align}
    y_t =& \space \,\exp(h_t/2) \varepsilon_t \\
    h_{t+1} =& \space \, \mu +\phi(h_t - \mu) + \sigma_{\eta} \eta_t  \\
    h_1 \sim& \space \,\mathrm{N}\left(\mu, \frac{\sigma_{\eta}^2}{1-\phi^2}\right) \\
    \varepsilon_t \sim& \space \,\mathrm{N}(0,1) \\
    \eta_t \sim& \space \,\mathrm{N}(0,1).
    \end{align}
    The measurement equation (1) describes the conditional distribution of $y_t$ given $h_t$, where $h_t$ is the conditional log variance of $y_t$. $h_1$ is a draw from a stationary distribution and the state equation $h_{t+1}$ follows a stationary process governed by the auto-regressive parameter $\phi$ such that $|\phi|<1$. This auto-regressive parameter represents the persistence or stickiness of log volatility and the dispersion parameter $\sigma_{\eta}$ is the constant variance of the states. $\epsilon_t$ and $\eta_t$ are standard normal white noise shocks and are uncorrelated with each other.

    KSC assign certain independent prior distributions and hyperparameters to the static parameters, with conjugate priors for $\mu$ and $\sigma^2$, given by
    \begin{align}
    \mu \sim& \space \,\mathrm{N}(0, 10) \\
    \sigma_{\eta}^2 \sim& \space \, \mathrm{IG}(shape=5/2, scale=(0.01\times 5) / 2) \\
    \phi^{\ast} \sim& \space \,\mathrm{Beta}(20, 1.5) \\
    \phi =& \space \, 2\phi^{\ast} - 1.
    \end{align}
    The prior distribution for $\phi$ is referred to as a stretched beta distribution, as it corresponds to a transformed beta random variable, $\phi^*$, with support (-1, 1).

@@

    Convergence diagnostics are useful for identifying when Bayesian computation fail on real data. However, confounding issues may arise when attempting to diagnose the cause of computational problems. Real data is generated from an unknown data generating process. That is, the true parameter and model are unobservable, and it is very likely that the model being fit is not the true model. Therefore, failed diagnostic checks could be the result of attempting to fit a misspecified model, issues with the MCMC algorithm, or both. Furthermore, different sampling strategies will result in different posterior estimates for the same model. Assuming no issues with the computation or the model specification, convergence diagnostics do not provide any guidance to the analyst regarding which posterior estimate is to be preferred.

    To illustrate the concerning issues, the posterior distribution associated with the stochastic volatility model in (1)-(5) and a set of real financial returns is estimated using two different MCMC algorithms. The data are demeaned daily (logarithmic) returns on the S\&P 500 Index over the period from 1 January 2023 - 10 September 2023 with sample size $T=171$. Figure \ref{fig:realdataex} displays the marginal posterior estimates for the static parameters $\mu$, $\phi$ and $\sigma^2$, with corresponding summary statistics in Table \ref{tab:realdata}. Overall the two methods are similar in skew and shape. However, closer inspections suggests that $\phi$ has a fatter left tail for the KSC sampler and $\sigma^2$ is also heavier in the right tails relative to HMC. This is evident in the quantiles with the 25th quantile for $\phi$ being smaller at 0.875 for KSC compared to 0.910 in HMC. Furthermore, HMC has taller peaks at their mode relative to KSC with a tighter spread. The uncertainty intervals created from each posterior sample would also be quite different and lead to different results. However, we can only conclude that there are differences between these methods, not which is correct.
        \begin{figure}[H]
        \centering
        \includegraphics[scale=0.1]{motivating_example/real_data_ex.png}
        \caption{\textbf{Posterior distribution for static parameters fit on the S\&P 500 Index.} Posterior samples are drawn from the HMC (blue) and KSC (red) MCMC algorithms. Marginal posterior densities are presented for $\mu$, $\phi$ and $\sigma^2$. Note that each panel has their own scale for the x and y axis.}
        \label{fig:realdataex}
    \end{figure}

    % Other methods for model selection in this context (for example, out of sample predictive performance), but for parameter estimation it is unclear which one should be selected. 
    \begin{table}[H]
        \centering
        \begin{tabular}{|c|c|c|c|c|c|c|c|} \hline 
        Parameter&  MCMC&Min& q25&  Median& Mean & q75&Max\\ \hline 
        $\mu$&  KSC&-2.03 & -0.472 & -0.354 & -0.355 & -0.231 & 1.31 \\
     $\mu$&  HMC&-4.55 & -0.508 & -0.373 & -0.370 & -0.229 &2.86  \\\hline 
     $\phi$&  KSC&0.384 & 0.875 & 0.928 & 0.902 & 0.958 & 0.997 \\
     $\phi$&  HMC&0.438 & 0.910 & 0.953 & 0.929 & 0.977 &1.00  \\ \hline 
     $\sigma^2$&  KSC&0.00237 & 0.00891 & 0.0138 & 0.0178 & 0.0210 & 0.130 \\ 
     $\sigma^2$&  HMC&0.00209 & 0.00731 & 0.0105 & 0.0131 & 0.0159 &0.107 \\ \hline
        \end{tabular}
        \caption{\textbf{Summary statistics for HMC and KSC algorithms}. Both sets of estimates appear reasonable but it is unclear from estimates on real data which estimate is closer to the truth.}
        \label{tab:realdata}
    \end{table}

\subsection{Challenge 2: Limitations of a single simulation}
    Diagnosing problems with the approximate posterior from models fit to real data is difficult because we do not observe the true data generating process or the true model. A strategy around this is to evaluate a model and algorithm on simulated data. One approach is to simulate data from a generative model using known parameters. Then fit the same model on the simulated data and see if the true parameters can be recovered. This gives us the benefit of defining the true parameters of the data generating process to be estimated. If the model and algorithm cannot adequately capture the true parameter, then we cannot be confident that it will provide reliable estimates on real data.

    Simulation is particularly important in the SV model as it has latent parameters. \citet{gelman2020bayesian} discuss that simulation is the only way we can check inference on latent variables. This is critical for the SV model since the underlying framework is a state space model with latent log variance parameters. Latent variables are unobserved in real data and are only estimated in the context of the model. Simulation gives control over the data generating process which reveals what the model can infer about the latent variables.

@@

    Figure 2 shows marginal posterior distributions given a single set of 1000 returns generated by the SV model. The true parameters are $\mu=-0.389$, $\sigma^2_{\eta}=0.0115$ and $\phi=0.547$ and the model is fit using Stan's implementation of HMC, the No-U-Turn sampler. The 95\% credible intervals for $\mu$ and $\sigma^2_{\eta}$ cover the true parameter. The true parameter for $\phi$ however, is in the tails and outside the interval. Such an analysis may incorrectly conclude that the model fails to adequately estimate the $\phi$ parameter. However, it may be the case that the posterior distribution is correctly calculated using this algorithm and the results may be due to the features of this specific simulated data set.

@@

\section{Methodology}
This section introduces the simulation design used to evaluate whether a MCMC algorithm is returning the correct posterior estimates on average (that is, whether a MCMC algorithm calibrated). Section 3.1 discusses the procedure for running simulation based calibration and Section 3.2 defines the diagnostics and metrics for evaluating SBC experiment results.\footnote{The software used to implement these methods and produce all results and plots in this paper are the R v4.2.3\citep{rlang}, Python v3.8.13 \citep{10.5555/1593511} and Stan v2.31.0 \citep{stan} programming languages, including open source R packages (tidyverse v1.3.2 \citep{tidyverse}, quantmod v0.4.20 \citep{quantmod}, cmdstanr 0.5.3 \citep{cmdstanr}, posterior v1.4.1 \citep{posteriorr}, bayesplot v1.4.10 \citep{bayesplot}, jsonlite v1.8.4 \citep{jsonlite}, arrow v12.0.1.1 \citep{arrow} and future.apply v1.11.0 \citep{RJ-2021-048}) and python packages (numpy v1.24.4 \citep{harris2020array}, scipy v.1.10.1 \citep{2020SciPy-NMeth}, statsmodels v.0.14.0 \citep{seabold2010statsmodels}, pandas v.2.0.3 \citep{mckinney-proc-scipy-2010} and pyarrow v12.0.1 \citep{arrow}).}

    \subsection{Simulation Design}
        Simulation Based Calibration (SBC) checks the calibration of posterior estimates generated by MCMC algorithms. SBC is conducted by comparing the distribution of rank statistics to the uniform distribution which arises when an algorithm is correctly calibrated. For each $k$ of $K$ SBC iterations, one value for each parameter is drawn from the prior. Conditional on the drawn parameter values, a data set is generated from the model. Then a sampling method (originally using Stan in \citet{talts2020validating}) is used to obtain posterior samples given this data and the prior. The estimated posteriors are then compared to the true values by calculating the rank statistics. Under a calibrated model, the $K$ rank statistics for each parameter should be uniformly distributed.

        To illustrate the SBC procedure, let $\theta$ be an arbitrary scalar parameter from a model and $y$ represent a data set. Start with a single parameter $\theta^{sim}$ drawn from the prior distribution
        \begin{align}
        \theta^{sim} \sim \pi(\theta),
        \end{align}
        then generate a data set $y^{sim}$ conditionally given $\theta^{sim}$ and the model of interest\footnote{Since all computation presumes a correctly specified model, notation to condition on the model is suppressed.}:
        \begin{align}
        y^{sim} \sim \pi (y|\theta^{sim}).
        \end{align}
        Take $L$ draws from the estimated posterior distribution, conditional on this data set, and generated by the MCMC algorithm under investigation (here, HMC or KSC):
        \begin{align}
        \{\theta_1,\dots , \theta_{B}\} \sim \pi (\theta | y^{sim}).
        \end{align}

        \citet{talts2020validating} prove that the sample of rank statistics for a given parameter, produced from replicate posteriors, follow a uniform distribution over the integers $\{0,1\dots,B\}$ if the posterior samples follow the same distribution as the prior on average. The rank statistic for any one-dimensional function of parameters $f:\Theta\rightarrow\ \mathbb{R}$ produced by the $k^{\mathrm{th}}$ replicate posterior sample, is defined as
        \begin{align}
        r^{(k)} = \mathrm{rank}(\{f(\theta_1^{(k)}),\dots , f(\theta_{B}^{(k)})\}, f(\theta^{sim})) = \sum_{b=1}^{B}1[f(\theta_{b}^{(k)}) < f(\theta^{sim})].
        \end{align}

@@

        A key result is that the posterior averaged over the data and true parameters will equal the prior distribution. This is shown in the following expression
        \begin{align}
        \pi(\theta) &= \int \int \pi(\theta|y^{sim}) \pi(y^{sim}|\theta^{sim}) \pi(\theta^{sim})d\theta^{sim} dy^{sim} \\
        &= \int \int \pi(\theta|y^{sim}) \pi(y^{sim},\theta^{sim}) d\theta^{sim} dy^{sim}.
        \end{align}
        That is, the average posterior for $\theta$ over the SBC iterations should approximately equal to prior distribution, if the algorithm is calibrated. Therefore, any deviation between the prior distribution and the average posterior draws produced from replicated posterior distributions means that the sampling methodology is not producing the correct posterior, on average.

@@

            The shape of the distribution of rank statistics gives insight into how an MCMC may be miscalibrated. Specifically, it gives information about how the average posterior deviates from the prior distribution of a given parameter. Figures \ref{fig:underestimation} and \ref{fig:underdispersed} are recreated from \citet{talts2020validating} for convenience. These display examples of posterior miscalibration and the corresponding non-uniform rank statistics.

@@

            Figure \ref{fig:underdispersed} has large inflated frequencies on both ends of the histogram resulting in a `U-shape'. \citet{talts2020validating} describe this as under-dispersion of the posterior relative to the prior distribution on average. The average posterior is too narrow (or overconfident) relative to the spread of the prior resulting in inflated frequency on both ends of the rank statistic distribution.

@@

            \begin{figure}[H]
                \centering
                \includegraphics[scale=0.07]{methodology/rhs.png}
                \caption{\textbf{Left:} Non uniform rank statistics with inflated frequency on the right side. \textbf{Right:} The average posterior distribution underestimates the prior distribution. The miscalibration from the rank statistic distribution comes from posterior samples underestimating the true parameter on average.}
                \label{fig:underestimation}
            \end{figure}

            \begin{figure}[H]
                \centering
                \includegraphics[scale=0.07]{methodology/underdispersed.png}
                \caption{\textbf{Left}: Non uniform rank statistics with inflated frequencies on both ends (`U-shape' distribution). \textbf{Right}: Average posterior is underdispersed relative to the prior distribution. Non-uniformity in rank statistics comes from posteriors which are over-confident relative to the prior.}
                \label{fig:underdispersed}
            \end{figure}

            \subsubsection{Chi squared statistics}
            One difficulty of relying on the visualisation of rank statistic distributions for SV models is there are a very large number of unknowns in the model to consider. An alternative to visually checking for uniformity of a given parameter is to calculate a Chi-squared statistic based on the counts in each histogram bin. Suppose the histogram has $J$ bins, where for each $j^{th}$ bin, $b_j$ is the number of counts and $e_j$ the expected count in bin $j$. The expected count is the ratio of $K$ rank statistics and $J$ bins, $e_j=\frac{K}{J}$. The corresponding Chi-squared statistic is denoted by
            \begin{align}
            \chi^2 = \sum_{j=1}^J \frac{(b_{j} - e_{j})^2}{e_j}.
            \end{align}
